Performing graph-based clustering with rOpenScPCA

### 2024-12-13 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
respond to reviews, and use rOpenScPCA:: throughout

### Input file

```{r paths}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
```

@@

```{r create process seurat, message = FALSE}
# Convert to a Seurat object
seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")

# Process the object with a Seurat pipeline to obtain clusters
seurat_obj <- seurat_obj |>
  SCTransform() |>
  RunPCA() |>
  FindNeighbors() |>
  FindClusters()
```

For the initial cluster calculations and evaluations, we will use the PCA matrix extracted from the SCE object.
It's also possible to use an SCE object or a Seurat object directly, which we will demonstrate later.
We'll show how to do this for both SCE and Seurat objects:

```{r extract pca data}
# Extract the PCA matrix from an SCE object
# Moving forward, we'll use this version for most examples
pca_matrix <- reducedDim(sce, "PCA")

# Extract the PCA matrix from a Seurat object
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
```

@@

```{r cluster sce}
# Calculate clusters with default parameters
cluster_results_df <- calculate_clusters(pca_matrix)

@@

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results_df <- calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity"
)

# Print resulting table
cluster_results_df
```

@@

```{r silhouette}
# calculate the silhouette width for each cell
silhouette_results <- calculate_silhouette(
  pca_matrix,
  cluster_results_df
)

# print result
silhouette_results
```

@@

```{r purity}
# calculate the neighborhood purity for each cell
purity_results <- calculate_purity(
  pca_matrix,
  cluster_results_df
)

# print result
purity_results
```

@@

```{r stability, warning=FALSE}
# calculate the stability of clusters
stability_results <- calculate_stability(
  pca_matrix,
  cluster_results_df
)

@@

```{r stability custom parameters}
# Calculate clusters with non-default parameters
cluster_df_leiden <- calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  pca_matrix,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

@@

If you are analyzing your data with a Seurat pipeline that includes calculating clusters, you can use `rOpenScPCA` to evaluate them.
We previously created a Seurat object called `seurat_obj` with clusters:

```{r show seurat clusters}
head(seurat_obj$seurat_clusters)
```

@@

# Create a data frame for input
seurat_cluster_df <- data.frame(
  cell_id = names(seurat_obj$seurat_clusters),
  cluster = seurat_obj$seurat_clusters
)
```

@@

```{r seurat silhouette}
seurat_silhouette_df <- calculate_silhouette(
  seurat_pca_matrix,
  seurat_cluster_df
)
seurat_silhouette_df
```

```{r seurat purity}
seurat_purity_df <- calculate_purity(
  seurat_pca_matrix,
  seurat_cluster_df
)
seurat_purity_df
```

@@

ScPCA cell metadata already contains a column called `cluster` with results from an automated clustering. 
These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses, with the following parameters:

* The Louvain algorithm
* Jaccard weighting
* 20 nearest neighbors

@@

```{r scpca silhouette}
# Calculate silhouette width
scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
scpca_silhouette_df
```

```{r scpca purity}
# Calculate neighborhood purity
scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
scpca_purity_df
```

When running `calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
We only recommend evaluating clusters with `calculate_stability()` if you know the original parameters used.

```{r scpca stability}
scpca_stability_df <- calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20
)
scpca_stability_df
```

@@

```{r run on sce}
# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(sce)
cluster_results_df
```

```{r run on seurat}
# Calculate clusters from a Seurat object using default parameters
cluster_results_df <- calculate_clusters(seurat_obj)
cluster_results_df
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
We can see this below with an SCE object, for example:

```{r use pc_name}
# First, we'll rename the PCA matrix for demonstration
reducedDimNames(sce) <- c("PCA_matrix", "UMAP")
reducedDimNames(sce)

# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(
  sce,
  pc_name = "PCA_matrix"
)
cluster_results_df
```

@@

Follow these steps save the clustering results to a Seurat object's cell metadata:

>>>>

```{r input file path}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")
```

@@

For the initial cluster calculations and evaluations, we will use the PCA matrix extracted from the SCE object.
It's also possible to use an SCE object or a Seurat object directly, which we will demonstrate later.

```{r extract pca data}
# Extract the PCA matrix from an SCE object
pca_matrix <- reducedDim(sce, "PCA")
```

@@

```{r cluster sce}
# Calculate clusters with default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(pca_matrix)

@@

```{r cluster sce nondefault}
# Calculate clusters with non-default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  nn = 15,
  objective_function = "modularity"
)

@@

```{r silhouette}
# calculate the silhouette width for each cell
silhouette_results <- rOpenScPCA::calculate_silhouette(
  pca_matrix,
  cluster_results_df
)

# Print the first rows of the resulting table
head(silhouette_results)
```

@@

```{r purity}
# calculate the neighborhood purity for each cell
purity_results <- rOpenScPCA::calculate_purity(
  pca_matrix,
  cluster_results_df
)

# Print the first rows of the resulting table
head(purity_results)
```

@@

```{r stability, warning=FALSE}
# calculate the stability of clusters
stability_results <- rOpenScPCA::calculate_stability(
  pca_matrix,
  cluster_results_df
)

@@

```{r stability custom parameters}
# Calculate clusters with non-default parameters
cluster_df_leiden <- rOpenScPCA::calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- rOpenScPCA::calculate_stability(
  pca_matrix,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

@@

If you are analyzing your data with a Seurat pipeline that includes calculating clusters, you can use `rOpenScPCA` to evaluate them.

To demonstrate this, we'll first create a Seurat object from our SCE object. 
We'll use the function `rOpenScPCA::sce_to_seurat()` for this, setting `use_symbols = FALSE` to preserve Ensembl ids as the feature names rather than converting to gene symbols.

```{r sce to seurat}
# Convert the SCE to a Seurat object using rOpenScPCA
seurat_obj <- rOpenScPCA::sce_to_seurat(sce, use_symbols = FALSE)

seurat_obj
```

The `rOpenScPCA::sce_to_seurat()` function will carry over the existing SCE column `cluster` into the Seurat object, which we'll use here for demonstration. 
If you calculate clusters within the Seurat ecosystem, you may need to use a different column name here, e.g. `seurat_clusters`.

```{r seurat cluster}
head(seurat_obj$cluster)
```

@@

# Create a data frame for input
seurat_cluster_df <- data.frame(
  cell_id = colnames(seurat_obj),
  cluster = seurat_obj$cluster
)
```

@@

```{r seurat silhouette}
seurat_silhouette_df <- rOpenScPCA::calculate_silhouette(
  seurat_pca_matrix,
  seurat_cluster_df
)
head(seurat_silhouette_df)
```

```{r seurat purity}
seurat_purity_df <- rOpenScPCA::calculate_purity(
  seurat_pca_matrix,
  seurat_cluster_df
)
head(seurat_purity_df)
```

@@

ScPCA cell metadata already contains a column called `cluster` with results from an automated clustering. 
These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses. 
We can see the parameters it used from the SCE object's metadata; all other parameters were left at their default values.

```{r extract cluster params}
# Print the clustering algorithm used
metadata(sce)$cluster_algorithm

# Print the weighting scheme
metadata(sce)$cluster_weighting

# Print the number of nearest neighbors
metadata(sce)$cluster_nn
```

@@

```{r scpca silhouette}
# Calculate silhouette width
scpca_silhouette_df <- rOpenScPCA::calculate_silhouette(
  pca_matrix,
  scpca_cluster_df
)
head(scpca_silhouette_df)
```

```{r scpca purity}
# Calculate neighborhood purity
scpca_purity_df <- rOpenScPCA::calculate_purity(
  pca_matrix,
  scpca_cluster_df
)
head(scpca_purity_df)
```

When running `rOpenScPCA::calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
We only recommend evaluating clusters with `rOpenScPCA::calculate_stability()` if you know the original parameters used.

```{r scpca stability}
scpca_stability_df <- rOpenScPCA::calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20
)
scpca_stability_df
```

@@

```{r run on sce}
# Calculate clusters from an SCE object using default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(sce)
cluster_results_df
```

```{r run on seurat}
# Calculate clusters from a Seurat object using default parameters
cluster_results_df <- rOpenScPCA::calculate_clusters(seurat_obj)
cluster_results_df
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.

@@

Follow these steps save the clustering results to a Seurat object's cell metadata.
Note that the Seurat function `AddMetaData()`, used below, assumes that the row order in the provided data frame matches the order of cells in the Seurat object you're adding metadata to.
